Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-25 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
comments on qEFT
footnote

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 1.33$~km
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{\hs{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field interaction arising from original the dilatonic coupling, i.e.,
$f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}.
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}.}}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{\rm f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond to 90\% credible values.
}
\label{tab:summary_qeft}
\end{table}

>>>>

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 1.33$~km
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{\hs{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field interaction arising from the original dilatonic coupling, i.e.,
$f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}.
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}.}}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{\rm f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond \hs{to 90\% credible values \emph{if} we allow for a more flexible cut-off $\varepsilon \lesssim 0.65$.
The result is however robust under the hard cut-off of $\varepsilon = 1/2$ at 65\% credible level.}
}
\label{tab:summary_qeft}
\end{table}
